A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-08-23 rigid_bodies.tex
Resumo em desenvolvimento

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}

\usepackage[pdftex]{graphicx}
%\usepackage{epstopdf}
%\usepackage{caption}
%\usepackage{subfig}
%\usepackage[]{units}
%\usepackage[]{natbib}
\usepackage[]{threeparttable}
%\usepackage{booktabs}
%\usepackage{multirow}

@@

\newenvironment{smallarray}[1]
{\null\,\vcenter\bgroup\scriptsize
	\renewcommand{\arraystretch}{0.7}
	\arraycolsep=.13885em
	\hbox\bgroup$\array{@{}#1@{}}}
{\endarray$\egroup\egroup\,\null}

@@

\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
\end{abstract}

@@

Improvements in molecular dynamics (MD) algorithms, along with advances in software and hardware, have allowed researchers to extend both the time and the length scales attainable in their atomistic simulations. This progress has fomented the ambitious attempt to elucidate complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few. Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.

One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion. This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles. However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach. It is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble. Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions. This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so on. It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.

In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as corresponding numerical integrators. The contribution in the NVE case relies on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom. We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions. For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and another one that corrects a pressure calculated in terms of interactions among free atoms.

The paper is organized as follows. In Sec.~\ref{sec:mechanics}, we present a mathematical reformulation of the Hamiltonian equations of motion for rigid bodies in terms of unit quaternions. In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVE and NVT ensembles, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Moreover, we present results related to kinetic energy partition in the NVE case, analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.

@@

In this section, we first review certain aspects of the kinematics of rigid bodies which are relevant to the present study. Unit quaternions are employed to represent the orientations of rigid bodies. Then, we introduce the use of a particular factorization of the rotation matrix, from which a mathematical relation arises between the rate of change of a four-dimensional quaternion and a three-dimensional angular velocity vector. An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal motion.

@@

As explained in the referred appendix, premultiplication of a vector with either $\mt B$ or $\mt C$ transforms it into a quaternion that lies in the hyperplane orthogonal to $\vt q$. Both transformations can be reversed by applying $\tr{\vt B}$ or $\tr{\vt C}$ accordingly. Therefore, one can define an intermediary frame of reference, which is fixed with respect to the unit quaternion $\vt q$ that represents the orientation of the body. Its connection to the body-fixed and space-fixed frames are summarized as follows:
\[
\boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\mt B}_{\tr{\mt B}}}
\boxed{\substack{\text{Quaternion-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\tr{\mt C}}_{\mt C}}
\boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
\]

Therefore, Eq.~\ref{eq:relation_qdot_omega} shows that $\dot{\vt q}$ is simply half the angular velocity vector expressed in the quaternion-fixed frame. Besides, it guarantees that the norm of $\vt q$ remains unchanged because Eq.~\ref{eq:diff_qTq} is identically satisfied.

@@

To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\citep{Goldstein2002} By carrying out the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
\begin{equation}
\label{eq:conj_momentum}
\vt \pi = 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega.
\end{equation}

@@

For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. The correct way of performing differentiation with respect to a constrained vector is described in Ref.~\onlinecite{Schay1995}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame. Therefore, $-\partial U/\partial \vt q$ is twice the torque represented in the quaternion-fixed frame. Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\tr{\mt B} \vt \pi = \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:euler_eq_quat}
\end{align}
\end{subequations}

@@

An exact solution of Eq.~\ref{eq:EDO_system_alternative} would be symplectic in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$. Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element. However, no closed-form solution is known, even in the simplest case of free motion (i.e., when $U$ is constant), due to the dependence of $\omega_k$ with $\vt q$ and $\vt \pi$ (Eq.~\ref{eq:omega_entry}) and the consequent coupling of Eqs.~\ref{eq:edo_q} and \ref{eq:edo_pi}. Nevertheless, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010}

For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
\[
\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
\]
where $i\!L$ is the Liouville operator. This equation has a formal solution expressed by $f(t) = e^{t i\!L}f_0$, where $f_0$ is the initial value of $f$. As $f$ can be a phase-space coordinate, the exponential operator $e^{t i\!L}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} represents a transformation from the initial condition to the phase-space point occupied at time $t$. Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.

@@

In this section, we present factorization schemes for generating molecular dynamics algorithms in the microcanonical (NVE) and canonical (NVT) ensembles. In both cases, a three-dimensional system is considered as consisting of $N$ interacting, non-collinear rigid bodies occupying a fixed volume $V$. An index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, including its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$). Thus, the phase space of the whole system now contains $3N$ spatial coordinates and $4N$ quaternion components, along with their conjugate momenta. Of course, $2N$ implicit constraints are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$.

@@

Considering the system described above, with $\vt r$ and $\vt q$ representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
\begin{equation}
\label{eq:H_NVE}
\mathcal{H} = U(\vt r,\vt q) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right).
\end{equation}

@@

Therefore, a symplectic integrator can be devised using the same splitting scheme as described in Sec.~\ref{sec:symplectic}. The only difference is that the action of each propagator (boost, free translation, and free uniaxial rotation) must now occur simultaneously for all rigid bodies in the system. If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\sum_{i=1}^N \vt p_i = \vt P$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\sum_{i=1}^N(\vt p_i t - m_i \vt r_i) = \boldsymbol{\mathcal G}$. If periodic boundary conditions are employed, as it is usual in molecular dynamics simulations, then there is no rotational symmetry and, consequently, no conservation of the total angular momentum.

@@

Kamberaj \textit{et al.}\cite{Kamberaj2005} proposed equations of motion and integration schemes for molecular dynamics with rigid bodies in the canonical ensemble...........

@@

Care must be taken when differentiating the potential energy $U$ with respect to $\vt q$ because the quaternion components are interdependent. Schay demonstrated in Ref.~\onlinecite{Schay1995} that, when differentiating a scalar function $f$ with respect to a vector variable $\vt x$, subject to a constraint $g(\vt x) = 0$, the derivative of $f$ with respect to $\vt x$ must be computed by
\[
\diff{f}{\vt x} = (\vt 1 - \vt n \tr{\vt n})\left( \diff{f}{\vt x} \right)^\ast,
\]
where $\vt n$ is the unit vector proportional to $(\partial g/\partial \vt x)^\ast$, with the asterisk denoting a differentiation performed without considering the interdependence of the entries of $\vt x$. We note that transpositions are necessary when comparing the equation above, based on the gradient convention, with that in Ref.~\onlinecite{Schay1995}, where the Jacobian convention is used. In the case of a unit quaternion,
\[
g(\vt q) = \tr{\vt q}\vt q - 1 = 0 \; \rightarrow \; \left(\diff{g}{\vt q}\right)^\ast = 2 \vt q.
\]

>>>>

\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}

\usepackage[pdftex]{graphicx}
\usepackage[]{threeparttable}

@@

\newenvironment{smallarray}[1]                          % small arrays
{\null\,\vcenter\bgroup\scriptsize
	\renewcommand{\arraystretch}{0.7}
	\arraycolsep=.13885em
	\hbox\bgroup$\array{@{}#1@{}}}
{\endarray$\egroup\egroup\,\null}

@@

\begin{abstract}
Sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or molecular substructures in Molecular Dynamics. This strategy eliminates degrees of freedom and admits larger time steps without giving up the atomistic character of a model. In this paper, we rely on a factorization of the rotation matrix, expressed in terms of unit quaternion components, to simplify the microcanonical and canonical dynamics of systems containing rigid bodies.
\end{abstract}

@@

Along with advances in software and hardware, improvements in molecular dynamics (MD) algorithms have allowed researchers to extend both the time and the length scales attainable in atomistic simulations. This progress has fomented the ambitious attempt of elucidating complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few. Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.

One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion. This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles. However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach. This is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble. Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions. This approach has been successfully employed in plain MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so forth. It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.

In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as numerical integrators for both cases. Our contribution in the NVE case resides on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom. We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions. For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.

The paper is organized as follows. In Sec.~\ref{sec:mechanics}, we present a mathematical reformulation of the Hamiltonian equations of motion for rigid bodies in terms of unit quaternions. In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVE and NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Moreover, we present results related to kinetic energy partition in the NVE case, analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.

@@

In this section, we first review certain aspects of the kinematics of rigid bodies which are relevant to the present study. Unit quaternions are employed to represent the orientations of rigid bodies. Then, we introduce the use of a particular factorization of the rotation matrix, from which a mathematical relation arises between the rate of change of a four-dimensional quaternion and a three-dimensional angular velocity vector. An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.

@@

As explained in the referred appendix, premultiplication of a vector with either $\mt B$ or $\mt C$ transforms it into a quaternion that lies in the hyperplane orthogonal to $\vt q$. Both transformations can be reversed by applying $\tr{\vt B}$ or $\tr{\vt C}$ accordingly. Therefore, there exists an intermediary frame of reference, which is fixed with respect to the unit quaternion $\vt q$ that represents the orientation of the body. Its connection to the body-fixed and space-fixed frames are summarized as follows:
\[
\boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\mt B}_{\tr{\mt B}}}
\boxed{\substack{\text{Quaternion-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\tr{\mt C}}_{\mt C}}
\boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
\]

Hence, Eq.~\ref{eq:relation_qdot_omega} shows that $\dot{\vt q}$ is simply half the angular velocity vector expressed in the quaternion-fixed frame. Besides, it guarantees that the norm of $\vt q$ remains unchanged because Eq.~\ref{eq:diff_qTq} is identically satisfied.

@@

To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\cite{Goldstein2002} By carrying out the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
\begin{equation}
\label{eq:conj_momentum}
\vt \pi = 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega.
\end{equation}

@@

For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. The correct way of performing differentiation with respect to a constrained vector is described in Ref.~\onlinecite{Schay1995}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame. Therefore, $-\partial U/\partial \vt q$ is twice the torque represented in quaternion-fixed frame of reference. Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\tr{\mt B} \vt \pi = \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:euler_eq_quat}
\end{align}
\end{subequations}

@@

An exact solution of Eq.~\ref{eq:EDO_system_alternative} would be symplectic in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$. Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element. Even though one cannot obtain a closed-form solution for Eq.~\ref{eq:EDO_system_alternative}, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010} For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
\[
\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
\]
where $i\!L$ is the Liouville operator. This equation has a formal solution expressed by $f(t) = e^{t i\!L}f_0$, where $f_0$ is the initial value of $f$. As $f$ can be a phase-space coordinate, the exponential operator $e^{t i\!L}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} represents a transformation from the initial condition to the phase-space point occupied at time $t$. Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.

@@

In this section, we present factorization schemes for generating molecular dynamics algorithms in the microcanonical (NVE) and canonical (NVT) ensembles. In both cases, a three-dimensional system is considered as consisting of $N$ interacting, non-collinear rigid bodies occupying a fixed volume $V$. An index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, including its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$). Thus, the phase space of the whole system now contains $3N$ spatial coordinates and $4N$ quaternion components, along with their $7N$ conjugate momenta. Of course, $2N$ implicit constraints are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$.

@@

Considering the system described above, with $\vt r$ and $\vt q$ now representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
\begin{equation}
\label{eq:H_NVE}
\mathcal{H} = U(\vt r,\vt q) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right).
\end{equation}

@@

Therefore, a symplectic integrator can be devised using the same splitting scheme of Sec.~\ref{sec:symplectic}, with the action of each propagator (boost, free translation, and free uniaxial rotation) occurring simultaneously for all rigid bodies in the system. If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\sum_{i=1}^N \vt p_i = \vt P$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\sum_{i=1}^N(\vt p_i t - m_i \vt r_i) = \boldsymbol{\mathcal G}$. If periodic boundary conditions are employed, as it is usual in molecular dynamics simulations, then there is no rotational symmetry and, consequently, no conservation of the total angular momentum.

@@

Kamberaj \textit{et al}.\cite{Kamberaj2005} proposed equations of motion and integration schemes for molecular dynamics with rigid bodies in the canonical ensemble...........

@@

Care must be taken when differentiating the potential energy $U$ with respect to $\vt q$ because the quaternion components are interdependent. Schay has demonstrated that, when a vector variable $\vt x$ is subject to a constraint $g(\vt x) = 0$, the derivative of $f$ with respect to $\vt x$ must be computed by\cite{Schay1995}
\[
\diff{f}{\vt x} = (\vt 1 - \vt n \tr{\vt n})\left( \diff{f}{\vt x} \right)^\ast,
\]
where $\vt n$ is the unit vector proportional to $(\partial g/\partial \vt x)^\ast$, with the asterisk denoting a differentiation performed without considering the interdependence of the entries of $\vt x$. We note that transpositions are necessary when comparing the equation above, based on the gradient convention, with that in Ref.~\onlinecite{Schay1995}, where the Jacobian convention is used. In the case of a unit quaternion,
\[
g(\vt q) = \tr{\vt q}\vt q - 1 = 0 \; \rightarrow \; \left(\diff{g}{\vt q}\right)^\ast = 2 \vt q.
\]
